---
title: "Entropy and multiscale complexity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Entropy and multiscale complexity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  dpi        = 150,
  out.width  = "100%"
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r packages, message = FALSE}
library(Rtractor)
```

## Three entropy measures, three different questions

Rtractor's entropy family answers three related but distinct questions:

- `perm_entropy()` -- how evenly distributed are the *ordinal patterns*
  (which of the last few points was biggest, second-biggest, and so on) in
  this series?
- `sample_entropy()` -- how likely is it that two short segments of this
  series, which match closely, continue to match one point further?
- `multiscale_entropy()` -- how does that same "likely to keep matching"
  question change as the series is progressively smoothed (coarse-grained)
  across a range of temporal scales?

The first two are single-scale, single-number summaries. The third is where
things get more interesting, and where the difference between "random" and
"complex" actually shows up.

## Sample entropy: template matching

`sample_entropy()` (Richman & Moorman 2000) compares every pair of
length-`m` templates in the series (excluding self-matches) within a fixed
Chebyshev-distance tolerance `sd(x) * r`, then does the same for length
`m + 1`, and reports `-log(matches at m+1 / matches at m)`. A series that's
mostly unpredictable has few long matches relative to short ones, so
`sample_entropy()` is high; a very regular series keeps matching, so it's
low:

```{r sample-entropy-basics}
set.seed(1)
white_noise  <- rnorm(1000)
smooth_signal <- sin(seq(0, 40 * pi, length.out = 1000))

sample_entropy(white_noise)
sample_entropy(smooth_signal)
```

## Why one number isn't enough: the multiscale idea

Here's the problem `multiscale_entropy()` (Costa, Goldberger & Peng 2002)
was built to solve. Compare two series with very different *structure* but
comparable single-scale entropy: pure white noise, and an autoregressive
(AR(1)) process with strong positive autocorrelation --the latter is
smoother and more predictable point-to-point, but has real structure that
plays out over multiple points, not just one:

```{r two-series}
set.seed(1)
white_noise       <- rnorm(2000)
correlated_signal <- as.numeric(
  stats::filter(rnorm(2000), filter = 0.9, method = "recursive")
)
```

`multiscale_entropy()` coarse-grains the series (non-overlapping
block-averaging) at each integer scale factor from `1` up to `scale_max`,
and computes `sample_entropy()` on each coarse-grained version -- crucially,
using a tolerance held *fixed* relative to the **original** series'
standard deviation at every scale, not each coarse-grained series' own
(shrinking) standard deviation. That's the detail that makes entropy values
comparable across scales at all; see `?multiscale_entropy`.

```{r mse-both}
mse_white <- multiscale_entropy(white_noise, scale_max = 10)
mse_corr  <- multiscale_entropy(correlated_signal, scale_max = 10)

round(mse_white$mse, 3)
round(mse_corr$mse, 3)
```

Look at the *shape* of these two profiles, not just the scale-1 value.
White noise's entropy decreases steadily as the scale grows -- coarse-
graining destroys the (already trivial) point-to-point unpredictability,
leaving less and less structure for `sample_entropy()` to find. The
correlated signal starts lower (it's more locally predictable than white
noise at scale 1) but its entropy *doesn't* decay the same way -- coarse-
graining doesn't destroy its structure nearly as fast, because that
structure plays out over several points rather than just one. By scale 5
or so, the correlated signal's entropy has overtaken white noise's.

```{r mse-plot, echo = FALSE, results = "asis"}
if (!has_ggplot2) {
  cat(
    "_ggplot2 isn't installed in the environment that built this article, ",
    "so the plot below is shown but wasn't executed._",
    sep = ""
  )
}
```

```{r mse-plot-real, eval = has_ggplot2, message = FALSE}
library(ggplot2)

mse_df <- data.frame(
  scale = c(mse_white$scale, mse_corr$scale),
  mse   = c(mse_white$mse, mse_corr$mse),
  series = rep(c("White noise", "Correlated (AR1, phi = 0.9)"), each = 10)
)

ggplot(mse_df, aes(scale, mse, colour = series)) +
  geom_line() +
  geom_point() +
  scale_colour_manual(values = unname(rtractor_palette("core")[c("steel_blue", "coral")])) +
  labs(
    title = "Multiscale entropy: white noise vs a correlated signal",
    subtitle = "Entropy decay with scale looks nothing alike, even though scale-1 values are similar",
    x = "Scale factor", y = "Sample entropy", colour = NULL
  ) +
  theme_rtractor()
```

This is the whole point of the multiscale approach: a single-scale entropy
value can't tell "genuinely unpredictable" apart from "structured across
multiple scales" -- you need the *profile* across scales to see the
difference. This is the same reasoning behind why physiological signals
(heart rate, EEG, gait) are usually characterised by their full MSE curve
rather than a single entropy number, in the original Costa et al. line of
work.

## Choosing `m` and `r`

Both `sample_entropy()` and `multiscale_entropy()` default to `m` = `2`,
`r` = `0.15` -- the standard convention from Costa et al.'s original MSE
papers, and a reasonable starting point for most physiological time series.
`m` is the template length being compared; `r` is the tolerance as a
fraction of the series' own standard deviation. Larger `r` tolerates more
noise before two templates stop "matching," which tends to reduce entropy
overall; smaller `m` needs less data to estimate reliably but captures less
structure. Changing either changes the absolute entropy values, so keep
them fixed across signals you intend to compare.

## References

Bandt C, Pompe B. Permutation entropy: a natural complexity measure for
time series. Phys Rev Lett 2002;88:174102.

Richman JS, Moorman JR. Physiological time-series analysis using
approximate entropy and sample entropy. Am J Physiol Heart Circ Physiol
2000;278(6):H2039-H2049.

Costa M, Goldberger AL, Peng CK. Multiscale entropy analysis of complex
physiologic time series. Phys Rev Lett 2002;89:068102.
