---
title: "Getting started with Rtractor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Rtractor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  dpi        = 150,
  out.width  = "100%"
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r packages, message = FALSE}
library(Rtractor)
```

## Overview

**Rtractor** is the complexity/statistical-physics layer of the Circadia Lab
R ecosystem: a shared home for the nonlinear dynamics and complex-systems
measures (entropy, fractal dimension, multifractal spectra, recurrence
quantification) that would otherwise get reimplemented piecemeal inside
signal-specific packages like `mrpheus`, `zeitR`, and `dynR`.

Like the rest of the ecosystem, Rtractor is **signal-agnostic**: every
function accepts a plain numeric vector regardless of where it came from --
EEG, actigraphy, BOLD, HRV, or anything else -- rather than assuming a
specific acquisition modality or staging scheme.

Where a solid reference implementation exists, Rtractor wraps it (via Rcpp)
rather than re-deriving the algorithm from scratch, to preserve numerical
parity with the original methods literature. Where no license permits a
direct wrap, functions are clean-room reimplementations from the published
algorithm, validated against the reference implementation on synthetic test
data. See `inst/COPYRIGHTS` for the full provenance of every function.

This article walks through everything currently implemented, organised by
family. See "What isn't here yet" at the end for what's still in progress.

## Installation

```r
# install.packages("remotes")
remotes::install_github("circadia-bio/Rtractor")
```

## Example data

A couple of synthetic series to work with throughout: white noise (an
uncorrelated signal, the classic complexity-measures benchmark) and a random
walk (its cumulative sum, the other classic benchmark).

```{r example-data}
set.seed(1)
white_noise <- rnorm(4000)
random_walk <- cumsum(white_noise)
```

## Fractal & multifractal analysis

### Detrended Fluctuation Analysis

`dfa()` estimates the scaling exponent alpha of a time series (Peng et al.
1994). By default it treats `x` as an increment series and integrates it
internally -- the standard DFA convention -- so white noise gives the
textbook benchmark of alpha ~ 0.5:

```{r dfa-white-noise}
dfa(white_noise)$alpha
```

Feeding a random walk through the same default pipeline amounts to *double*
integration -- the other classic benchmark, alpha ~ 1.5:

```{r dfa-random-walk}
dfa(random_walk)$alpha
```

### Higuchi Fractal Dimension

`higuchi_fd()` estimates fractal dimension from curve length at increasing
sub-sampling intervals (Higuchi 1988). White noise is space-filling
(HFD ~ 2); a smooth periodic signal is line-like (HFD ~ 1):

```{r higuchi}
higuchi_fd(white_noise, k_max = 10)$hfd

smooth_signal <- sin(seq(0, 40 * pi, length.out = 4000))
higuchi_fd(smooth_signal, k_max = 10)$hfd
```

### Multifractal Detrending Moving Average (MFDMA)

`mfdma()` extends DFA to a spectrum of scaling exponents across multifractal
orders `q` (Gu & Zhou 2010), returning the singularity spectrum f(alpha):

```{r mfdma, fig.height = 3.5}
mf <- mfdma(white_noise, n_min = 10, n_max = 400, n_scales = 20)
plot(mf$alpha, mf$f, type = "b", xlab = "alpha", ylab = "f(alpha)")
```

White noise is close to monofractal, so the spectrum collapses to a narrow
range around alpha ~ 0.5 with f(alpha) peaking near 1.

### Chhabra-Jensen multifractal spectrum

`chhabra_jensen()` estimates the same kind of spectrum via direct
box-counting (Chhabra & Jensen 1989) rather than detrended fluctuations. It
needs a strictly positive series with a dyadic (power-of-two-friendly)
length:

```{r chhabra-jensen, fig.height = 3.5}
positive_series <- abs(rnorm(1024)) + 0.01
cj <- chhabra_jensen(positive_series, scales = 1:6)
plot(cj$alpha, cj$falpha, type = "b", xlab = "alpha", ylab = "f(alpha)")
```

Each `q` value's `alpha`/`falpha`/`Dq` estimate comes with an R-squared
(`r_squared_alpha`, `r_squared_falpha`, `r_squared_Dq`) -- worth checking
before trusting any individual point, especially near the edges of the `q`
range.

## Nonlinear time-domain features

Three fast, cheap-to-compute descriptors centralised from `mrpheus`'s AASM
staging feature pipeline (itself a validated port of the `antropy`/YASA
Python feature set):

```{r nonlinear-features}
petrosian_fd(white_noise)

hjorth_parameters(white_noise)

num_zerocross(white_noise)
```

`petrosian_fd()` is a fast proxy for irregularity based on sign changes in
the first difference. `hjorth_parameters()` returns mobility (a proxy for
mean frequency) and complexity (a proxy for bandwidth) from variance ratios
of successive differences -- note this uses Bessel-corrected variance,
matching R's `var()` convention, which differs slightly from `antropy`'s
population-variance convention for short series (see `?hjorth_parameters`).
`num_zerocross()` simply counts sign changes.

## Entropy

`perm_entropy()` estimates complexity from the distribution of ordinal
patterns in the series (Bandt & Pompe 2002), normalised to `[0, 1]` by
default:

```{r perm-entropy}
perm_entropy(white_noise)
perm_entropy(smooth_signal)
```

White noise has near-maximal permutation entropy (every ordinal pattern is
close to equally likely); the smooth periodic signal has much lower entropy
(a small number of patterns dominate).

## Recurrence quantification analysis (RQA)

`recurrence_microstate_entropy()` implements a parameter-free approach to
recurrence analysis (Corso et al. 2018): rather than picking a vicinity
threshold epsilon by hand, it searches for the threshold that *maximises*
the Shannon entropy of the recurrence microstate distribution.

```{r microstates}
windowed_signal <- (sin(seq(0, 20 * pi, length.out = 300)) + 1) / 2
recurrence_microstate_entropy(windowed_signal, seed = 1)
```

## Colour palette & theme

Rtractor ships its own colour palette and `ggplot2` theme, visually distinct
from `circadia`'s (softer, more pastel) so figures from each package are
recognisable at a glance:

```{r palette}
rtractor_palette()
rtractor_palette("core")
```

```{r check-ggplot2, echo = FALSE, results = "asis"}
if (!has_ggplot2) {
  cat(
    "_ggplot2 isn't installed in the environment that built this article, ",
    "so the plot below is shown but wasn't executed._",
    sep = ""
  )
}
```

```{r theme-demo, eval = has_ggplot2, message = FALSE}
library(ggplot2)

mf_df <- data.frame(alpha = mf$alpha, f = mf$f)

ggplot(mf_df, aes(alpha, f)) +
  geom_point(colour = rtractor_palette("core")[["steel_blue"]], size = 2) +
  geom_line(colour = rtractor_palette("core")[["steel_blue"]]) +
  labs(
    title = "MFDMA singularity spectrum",
    subtitle = "White noise: close to monofractal",
    x = expression(alpha), y = expression(f(alpha))
  ) +
  theme_rtractor()
```

## What isn't here yet

Several planned families aren't implemented yet:

- **Lyapunov exponents** (`R/lyapunov.R`) -- Rosenstein and Wolf methods for
  the largest Lyapunov exponent.
- **Multiscale metrics** (`R/multiscale.R`) -- multiscale entropy and
  refined composite multiscale entropy, built on the entropy family applied
  at each coarse-grained scale.
- **General RQA measures** (`R/rqa.R`) -- the recurrence matrix itself and
  its derived quantifiers (determinism, laminarity, recurrence rate,
  trapping time). `recurrence_microstate_entropy()` is a threshold-selection
  tool, not a replacement for these.
- **Phase-space embedding** (`R/embed.R`) -- time-delay embedding, and
  delay/dimension estimation, needed by the Lyapunov and RQA families above.

See `NEWS.md` for progress, or the package's GitHub repository for the
current status of reference code for each.
