---
title: "Multifractal methods: MFDMA vs Chhabra-Jensen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multifractal methods: MFDMA vs Chhabra-Jensen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  dpi        = 150,
  out.width  = "100%"
)
```

```{r packages, message = FALSE}
library(Rtractor)
```

## Two estimators, one underlying idea

Rtractor ships two multifractal spectrum estimators: `mfdma()` (Gu & Zhou
2010) and `chhabra_jensen()` (Chhabra & Jensen 1989). Both try to answer
the same question -- does this time series need more than one scaling
exponent to describe it, and if so, what's the full spectrum of exponents
present? -- but they get there differently.

`mfdma()` is a multifractal generalisation of DFA: it detrends the series
against a moving average at a range of scales, then looks at how the q-th
order fluctuation function scales with segment size. `chhabra_jensen()`
instead treats the series as a measure and does direct box-counting,
avoiding the Legendre transform that `mfdma()` (and MF-DFA) rely on.

That difference matters in practice. França et al. (2018) benchmarked
MF-DFA, MF-DMA, and Chhabra-Jensen against each other on both simulated and
human intracranial EEG data, and found Chhabra-Jensen the most stable of
the three across repeated epochs of the same underlying signal -- the
reason it's implemented here alongside `mfdma()`, rather than reaching for
MF-DFA. See `?chhabra_jensen` for the full reference.

## Ground truth, not just plausible-looking data

It's easy to run a multifractal estimator on `rnorm()` and see *some*
output, but that doesn't tell you whether the estimator is actually
detecting multifractality correctly -- white noise is close to monofractal,
so a badly-behaved estimator could still produce a narrow-looking spectrum
purely by chance.

`pmodel()` solves this by generating a p-model binomial cascade (Meneveau &
Sreenivasan 1987) whose multifractal properties are directly controlled by
its `p` parameter: values near `0.5` are essentially monofractal, and values
far from `0.5` are strongly multifractal. This gives both estimators a
signal with **known** ground truth to be checked against, rather than just
a signal that happens to look complex.

```{r pmodel-series}
y_calm   <- pmodel(8192, p = 0.48, seed = 1)  # near-monofractal
y_multi  <- pmodel(8192, p = 0.15, seed = 1)  # multifractal
y_strong <- pmodel(8192, p = 0.05, seed = 1)  # strongly multifractal
```

Both estimators expect the *raw* p-model output directly -- `mfdma()` and
`chhabra_jensen()` each do their own internal integration/box-counting, so
don't difference or log-transform it first.

## MFDMA on the ground-truth series

```{r mfdma-ground-truth}
mf_calm   <- mfdma(y_calm,   n_min = 10, n_max = 400, n_scales = 25)
mf_multi  <- mfdma(y_multi,  n_min = 10, n_max = 400, n_scales = 25)
mf_strong <- mfdma(y_strong, n_min = 10, n_max = 400, n_scales = 25)

widths_mfdma <- c(
  calm   = diff(range(mf_calm$alpha)),
  multi  = diff(range(mf_multi$alpha)),
  strong = diff(range(mf_strong$alpha))
)
widths_mfdma
```

The estimated spectrum width increases monotonically as `p` moves away from
`0.5`, exactly as the underlying cascade predicts -- `mfdma()` is genuinely
picking up the multifractal structure, not just returning a wide spectrum
regardless of input.

## Chhabra-Jensen on the same series

```{r cj-ground-truth}
cj_calm   <- chhabra_jensen(y_calm,   scales = 1:11)
cj_multi  <- chhabra_jensen(y_multi,  scales = 1:11)
cj_strong <- chhabra_jensen(y_strong, scales = 1:11)

widths_cj <- c(
  calm   = diff(range(cj_calm$alpha)),
  multi  = diff(range(cj_multi$alpha)),
  strong = diff(range(cj_strong$alpha))
)
widths_cj
```

Same pattern, and the two estimators land within a few percent of each
other on the same data -- independent cross-validation, not just each
method being internally consistent with itself.

```{r comparison-plot, fig.height = 3.8}
plot(
  mf_multi$alpha, mf_multi$f, type = "b", col = rtractor_palette("core")[["steel_blue"]],
  xlab = expression(alpha), ylab = expression(f(alpha)),
  main = "p = 0.15: MFDMA vs Chhabra-Jensen",
  ylim = c(0, 1.05)
)
lines(cj_multi$alpha, cj_multi$falpha, type = "b", col = rtractor_palette("core")[["coral"]])
legend(
  "bottomright", legend = c("MFDMA", "Chhabra-Jensen"),
  col = rtractor_palette("core")[c("steel_blue", "coral")], lty = 1, pch = 1, bty = "n"
)
```

`chhabra_jensen()`'s spectrum also has a theoretical constraint worth
knowing about: the peak height, `max(falpha)`, should sit close to `1` --
the fractal dimension of the one-dimensional support the measure lives on.
That's a useful sanity check independent of the p-model comparison above:

```{r peak-check}
max(cj_multi$falpha)
```

## Which one should I use?

- If you want the more *stable* estimate across repeated epochs of the same
  signal, and your series can be made strictly positive (or you're willing
  to apply a sigmoid transform), reach for `chhabra_jensen()` first -- see
  França et al. (2018) for why.
- If your series already has a natural DFA-style workflow around it, or you
  need `mfdma()`'s cross-package parity with other MF-DFA-family tools,
  `mfdma()` is the more familiar entry point.
- Either way, check the R-squared / fit diagnostics each function returns
  (`r_squared_alpha`, `r_squared_falpha`, `r_squared_Dq` for
  `chhabra_jensen()`) before trusting an individual `q` value's estimate,
  especially near the edges of the `q` range.

## References

Chhabra A, Jensen RV. Direct determination of the f(alpha) singularity
spectrum. Phys Rev Lett 1989;62:1327-1330.

Gu GF, Zhou WX. Detrending moving average algorithm for multifractals.
Phys Rev E 2010;82:011136.

Meneveau C, Sreenivasan KR. Simple multifractal cascade model for fully
developed turbulence. Phys Rev Lett 1987;59:1424-1427.

Franca LGS, Miranda JGV, Leite M, Sharma NK, Walker MC, Lemieux L, Wang Y.
Fractal and multifractal properties of electrographic recordings of human
brain activity: toward its use as a signal feature for machine learning in
clinical applications. Front Physiol 2018;9:1767.
